A content-based model of how experiences and memories unfold over time reveals event-like structure in naturalistic episodic recall

### 2019-06-18 paper/main.tex
saving progress on discussion

\newcommand{\topicopt}{S1}
\newcommand{\topics}{S2}
\newcommand{\featureimportance}{S3}
\newcommand{\listlearning}{S4}
\newcommand{\corrmats}{S5}
\newcommand{\matchmats}{S6}
\newcommand{\wasserstein}{S7}

@@

Statistical models of memory studies often treat memory recalls as "all-or nothing" and independent events.  However, our approach produces a content-based model of individual stimulus events and recall events, allowing for direct quantitative comparison between all stimulus and recall events, as well as between the recall events themselves.  Leveraging this affordance, we introduce 2 novel metrics for quantifying memory representations: precision and distinctiveness.  We define precision as the average correlation between each recall event and the maximally correlated video event [INSERT FIGURE REF].  Participants whose recall events are more veridical descriptions of what happened in the video event will have higher precision scores. We find that across participants, a higher precision score is correlated to both hand annotated memory performance as well as the number of recall events (per participant) estimated by our model [INSERT STATS, INSERT FIGURE REF]. A second novel metric we introduce here is the distinctiveness, or the uniqueness of each recall event relative to other recall events. Distinctiveness is defined as 1 minus the average of the upper triangle of the recall-by-recall correlation matrix. Similar to precision, we find that the more distinct participants recalls are (on average), the more they remembered (for both hand-annotated and model derived memory performance) [INSERT STATS AND FIGURE REF].  In summary, using two novel metrics afforded by our approach, we find that participants whose recalls are both more precise and distinct remember more content.

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_distinctiveness}
  \caption{\small \textbf{Novel content-based metrics of naturalistic memory: precision and distinctiveness.} \textbf{A.} Blah-bahdy Blah}
  \label{fig:precision-distinctiveness}
\end{figure}

@@

In this work, we used topic models to capure the contents of verbal recall.  While a large number of language models exist (e.g. WAS, LSA, word2vec), we

>>>>

\newcommand{\topicopt}{S1}
\newcommand{\topics}{S2}
\newcommand{\featureimportance}{S3}
% \newcommand{\listlearning}{S4}
\newcommand{\corrmats}{S4}
\newcommand{\matchmats}{S5}
\newcommand{\wasserstein}{S6}

@@

Statistical models of memory studies often treat memory recalls as binary (e.g. the item was recalled or not) and independent events.  However, our approach produces a content-based model of individual stimulus events and recall events, allowing for direct quantitative comparison between all stimulus and recall events, as well as between the recall events themselves.  Leveraging this affordance, we introduce 2 novel metrics for quantifying memory representations: precision and distinctiveness.  We define precision as the average correlation between each recall event and the maximally correlated video event (Fig.~\ref{fig:precision-distinctiveness}).  Participants whose recall events are more veridical descriptions of what happened in the video event will presumably have higher precision scores. We find that across participants, a higher precision score is correlated to both hand annotated memory performance (Pearson's $r(15) = 0.6, p = 0.011$) as well as the number of recall events (per participant) estimated by our model (Pearson's $r(15) = 0.64, p = 0.005$). A second novel metric we introduce here is the distinctiveness, or the uniqueness of each recall event relative to other recall events. Distinctiveness is defined as 1 minus the average of the upper triangle of the recall-by-recall correlation matrix. Similar to precision, we find that the more distinct participants recalls are (on average), the more they remembered (hand-annotated memory: Pearson's $r(15) = 0.83, p < 0.001$ and model derived memory: Pearson's $r(15) = 0.71, p = 0.001$).  In summary, using two novel metrics afforded by our approach, we find that participants whose recalls are both more precise and distinct remember more content.

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_distinctiveness}
  \caption{\small \textbf{Novel content-based metrics of naturalistic memory: precision and distinctiveness.} \textbf{A.} A video-recall correlation matrix for a representative participant (17).  The yellow boxes highlight the maximum correlation in each column.  Precision was computed as the average of the maximum correlation in each column.  On the other hand, distinctiveness was defined as the average of everything except for the maximum correlation in each column. \textbf{B.} The (Pearson's) correlation between precision and hand-annotated memory performance. \textbf{C.} The correlation between precision and the number of events recovered by the model ($k$). \textbf{D.} The correlation between distinctiveness and hand-annotated memory performance. \textbf{E.} The correlation between distinctiveness and the number of events recovered by the model ($k$).}
  \label{fig:precision-distinctiveness}
\end{figure}

@@

Many statistical models of memory treat recall as binary and unrelated events. Our novel framework allows one to measure the \textit{precision} of each recall event (by considering the correlation to the highest video event) as well as how \textit{distinctive} each recall event is from other recall events.  These new metrics afforded by our method allow for much more nuanced analysis of recall dynamics than were previously possible.  Using them, we find that both precision and distinctiveness are correlated to the amount of information participants recall. Thus, the manner in which a person recounts a \textit{particular} event is related to their ability to recall \textit{many} events.

While a large number of language models exist (e.g. WAS, LSA, word2vec, universal sentence encoder), here we use topic models for a few reasons.  First, topic models capture the \textit{essence} of a text passage devoid of the specific set and order of words used.  This was an important feature of our model since different people may accurately recall a scene using very different language. Secondly, words can mean different things in different contexts (e.g. baseball bat vs. the animal bat), and topic models are robust to this since words can be part of multiple topics.  Lastly, topic models provide a straight forward to recover the weights for the particular words comprising a topic, allowing for easy interpretation of an event's contents (e.g. Fig.~\ref{fig:topics}). Other models such as Google's universal sentence encoder offer a context-sensitive encoding of text passages, but the encoding space is complex and non-linear and thus, recovering the original words used to fit the model is not straight forward.

However, it's worth pointing out that our framework is divorced from the particular choice of language model. Moreover, many of the aspects of our framework could be swapped out for other choices. For example, the language model, the timeseries segmentation model and the video-recall matching function could all be customized for the particular problem. Indeed for some problems, recovery of the particular recall words may not be necessary, and thus other text-modeling approaches (such as universal sentence encoder) may be preferable. Future work will explore the influence of particular model choices on the framework's accuracy.
